Subset coercions in \Coq \\ \small{Draft - \today}

### 2006-09-01 papers/article/main.tex
New notation for sub args

The option type is the usual way to encode partiality in
\Coq. However we would more naturally constrain the argument \id{l} to
non-empty lists to be more faithful to the original \ML code, and also
attach a property to the result to express more precisely what \id{tail}
does:
\begin{code}
  Definition tail (l : list A) : forall p : l <> [] -> 
  \{ l' : list A | exists a, a :: l' = l \}.
\end{code}
Now the definition's body requires some heavy plumbing of the code which is 
not affordable when programming. Moreover, the user is forced to give a
proof term when calling the function.
We propose a solution to overcome these difficulties, based on the \ps{}
\cite{Shankar&Owre:WADT99} technique of \PVS{}
\cite{PVS-Semantics:TR}. It is a weak, \emph{decidable} type-checking
procedure which does not require proofs to be present in the code when
constructing objects of a subset type. Using our new language \Russell, we would write:
\begin{code}
  Definition tail \{ l : list A | l <> [] \} : 
    \{ l' : list A | exists a, a :: l' = l \} := 
  match l with
    hd :: tl -> tl
  | [] -> []
  end
\end{code}

The specification shows that we're defining a partial function and
shows the relation between input and output, using a dependent type, yet the
code remains as simple as the \ML{} definition. This is only possible
because we leave the proofs in both branches. After type-checking,
there is an automatic elaboration into partial \Coq terms, which collects
obligations the user has to prove. In our example,
the proof assistant will ask the user to prove that:
\begin{enumerate}
\item The list \verb|tl| has the property \verb|exists a, a :: tl = hd :: tl|
  and
\item In the context where
  \id{l} is a non-empty list and $\texttt{l} = []$, the list $[]$ has the property
  \verb|exists a, a :: [] = []| (which should be obvious as the context is contradictory).
\end{enumerate}

@@

In \Coq, the distinction between informative and propositional parts of
a term is formalized by the \Prop/\Set sorts. In \Russell, we have
special support for propositions appearing in subset types. By
delimiting the use of propositions, we can separate code from proof.

Informally, this idea is already present in mathematics. When you have a
subset $\{ x `: S `| P \}$, you can freely forget about the property $P$
and use any operation which is defined on $S$. Conversely, when you
want to use an operation defined on a subset, say $f : \{ x : `N `| P \} "->"
X$, you usually prove \emph{first} that you apply it only to elements 
having the desired property. For example $(f~2)$ is a correct application
only if $(P~2)$ is provable.
In the context of formal development of programs, such a workflow is not
entirely satisfactory because it forces one to create objects and prove
properties about them \emph{at the same time}. In \Coq{}, we are forced
to apply $f$ to an object $(\elts{2}{p})$ where $p$ is a proof of
$(P~2)$ (figure \ref{fig:cci-subset} presents the definition of the subset type in \Coq).
We would like to be able to prove that our usage of partial functions 
is correct only after the program is written. Subsets are particularly well-suited in this respect
because they separate the objects we want to manipulate and their associated
properties.

\begin{figure}[h]
  \begin{center}
    \def\type{\typed}
    \AXC{$`G \type A : \Set$}
    \AXC{$`G, x : A \type P : \Prop$}
    \BIC{$`G \type \mysubset{x}{A}{P} : \Set$}
    \DP

@@

    \AXC{$`G \type a : A$}
    \AXC{$`G \type p : P[a/x]$}
    \AXC{$`G \type \mysubset{x}{A}{P} : \Set$}
    \TIC{$`G \type \elt{A}{(\lambda x : A.P)}{a}{p} : \mysubset{x}{A}{P}$}
    \DP
    \vspace{\infvspace}

    \AXC{$`G \type t : \mysubset{x}{A}{P}$}
    \UIC{$`G \type \eltpit~t : A$}
    \DP
    \quad
    \AXC{$`G \type t : \mysubset{x}{A}{P}$}
    \UIC{$`G \type \eltpip~t : P[\eltpit~t/x]$}
    \DP
  \end{center}
  \vspace{-2em}
  \caption{Subset type in \CIC}
  \label{fig:cci-subset}
  \vspace{-1em}
\end{figure}

@@

%
\Russell{}'s type system is based on the Calculus of (Co-)Inductive
Constructions (figure \vref{fig:cic-decl-rules}) \cite{coquandhuet88},
with sigma types but without universes.  This restriction may be removed
in future work, but cause no problem for our main purpose which is
programming. We omit (co-)inductive constructs here, as they can be
considered as constants and leave their treatment as future work. The
judgement $`G \typed t : T$ means $t$ is a well-typed term of type $T$
in environment $`G$.  In \Russell, the conversion rule is replaced by a
new subsumption rule \irule{Subsum} (figure
\vref{fig:typing-decl-rules}) which will implement the subset
equivalence. The judgement $`G \typed T \subd U : s$ means $T$ is
equivalent to $U$ in environment $`G$, both sorted with $s$.
Following the presentation as a Pure Type System, the set of sorts
$\PTSsorts$ is defined as $\setproptype$. As usual, we let $s, s_i$ for
$i `: `N$ range over sorts. The axioms are $\PTSaxioms = \{ (\Set,
\Type), (\Prop, \Type) \}$ and the set of rules $\PTSrules$ is defined
by the functional relation $`A s_1 s_2, (s_1, s_2, s_2) `: \PTSrules$. We allow products $\Pi x :
A, B$ where $A : \Prop$ and $B : \Set$ but the user is encouraged to
encode them as $\Pi x : \{ \_ : unit `| A \}, B$ products to benefit from
the subtyping mechanism.

@@

Likewise, we need to restrict the system so that it becomes property
irrelevant for the propositions appearing in subset types. Indeed, in
\PVS{} you can't manipulate proof terms, so you have no way to create a
term of type $0 \neq 0$ from the judgement $0 : \mysubset{x}{`N}{x \neq
  0}$, nor can you actually derive the latter, as it won't type-check.
However, in a classical Calculus of Constructions with sigma types or inductive
types, the subset type is a first class object defined with proper
introduction and elimination terms (figure \vref{fig:cci-subset}). These
are not compatible with our subtyping, so we need to forbid the use of
these terms in \Russell{} terms, leaving only the type former (figure
\vref{fig:typing-decl-rules}). With this restriction, when we erase subset types
from \Russell terms, leaving only the supports, we get a valid
\CC term, hence $`_$ is not provable in \Russell.

We now need to change the conversion rule to a stronger one which will
include our subset equivalence.

@@

\begin{paragraph}{}
We use a standard transformation to get an algorithm from our typing
rules, integrating the conversion (or subsumption) rule
(here \irule{Coerce}) to the premises of the other rules.
%We recall that $`G \type t : s$ is shorthand for $`G \type t : T$ and
%$T  `= s$ with $s `: \PTSsorts$ in a algorithmic presentation.
Our algorithmic judgement $`G \typea t : T$
(figure \vref{fig:typing-algo-rules})
reads: $t$ is a well-typed term of type $T$ in environment $`G$.
We need to introduce the notion of support for subset types in order to
define our algorithmic system. Indeed, when typing an application, 
we need to ensure that the object we apply can be seen as an object of a
product type modulo the equivalence. 
% This notion is presented in \PVS{} by
% means of a $\mu_0$ operator which was readily adaptable in our setting
% (figure \vref{fig:mualgo-definition}). In Gang Chen's work, the notion is formalized by a
% new judgement $\Pi_{lub}$ which is essentialy equivalent to the operator
% $\mualgo$, except ours finds $\Sigma_{lub}$s too.
\begin{figure*}[t]
  \def\infvspace{0.5em}
    \def\type{\typea}
    \def\subt{\subta}
    \def\sub{\suba}
    \def\fCenter{\typea}
    \begin{center}
%     \WfEmpty\DP

@@

\end{paragraph}
\begin{paragraph}{}
  The next thing to do is to construct a decidable judgement for
  equivalence given two types. We denote it by $`G \typea T
  \suba U : s$ (figure \ref{fig:subtyping-algo-rules}).
  In our case, we have to check that conversion is decidable, which is the
  case. We also need to eliminate the transitivity and symmetry rules which are not
  syntax-directed. That is the purpose of the following theorem:
\begin{theorem}[Admissibility of transitivity and symmetry]
  \begin{enumerate}
  \item If $`G \typea T \suba U : s$ and $`G \typea U \suba V : s$, then $`G \typea T
    \suba V : s$.
  \item If $`G \typea T \suba U : s$ then $`G \typea U \suba T : s$.
  \end{enumerate}
\end{theorem}
\begin{proof}
  \begin{enumerate}
  \item By induction on the sum of the depths of the two derivations.
  \item Immediate using a thinning lemma.
  \end{enumerate}
\end{proof}
\end{paragraph}

@@

\begin{theorem}[Completeness]
  If $`G \typed t : T$ then there exists $T'$, $`G \typea t : T'$ and $`G \typea T' \suba T : s$.
  If $`G \typed T \subd U : s$ then $`G \typea T \suba U : s$.
\end{theorem}

@@

We denote by $\typeafn{`G}{t}$ the function which returns the type of
term $t$ in context $`G$.

@@

We can now prove the following:
\begin{theorem}[Forgetful map correctness]
  If $`G \typec t : T$ then $`G^{\circ} \typea t^{\circ} : T^{\circ}$ if $()^{\circ}$ is
  defined (ie, does not return terms containing $`_$) on $`G, t$ and $T$.
\end{theorem}

@@

We describe the case of application here, others are similar. First we
get the types $F$ and $U$ of the function $f$ and argument $u$ by recursively calling the
typing function on both terms. Then we must ensure that $F$ can indeed
be seen as a product using the $\mualgo$ operator. We can then build
the coercion between $F$ and this product, and between $U$ and its
domain. We finally instantiate these coercions by their corresponding
interpreted objects and return their application.

@@

Similarly, we must add $\eta$ rules for all introduction terms of our
language if we want to be conservative and have unicity of
coercions. The later is an important property of our system, as it
corresponds in practice to the determinism of proof-obligations
generation. The complete equational theory of \CICq{} is given in figure
\vref{fig:eqccq}. It includes $\eta$, surjective pairing for dependent
sums and subsets ($\rho$) and proof-irrelevance for the second component
of subset objects.
Benjamin Werner has studied the addition of proof-irrelevance (rule
$\sigma$) in the calculus of constructions
\cite{Werner:ProofIrrelevance}. It is not a trivial extension and it has
far-reaching consequences on the model of the calculus
\cite{DBLP:conf/types/MiquelW02}. We direct the reader to these papers for further
information on the subject.
\begin{figure}[ht]
  \vspace{-1em}
  \ccqeqarr
  \vspace{-2em}
  \caption{Equational theory of \CICq}
  \label{fig:eqccq}
  \vspace{-1em}
\end{figure}

@@

\section{Implementation}
\label{sec:Program}
This mechanism has been implemented in \Coq and tested on simple
examples. We actually generate coercions simultaneously with typing 
in the implementation, which is sound thanks to the previous proofs. Our
typing is a clone of \Coq's original typing algorithm, so we benefit
from all the features of \Coq including implicit variables, notations and the existing
coercion system. The type-checker also transforms pattern-matching so
that an equality between the matched term and the pattern is present in
each branch typing context and it allows the user to put explicit holes in the term, for example:
\begin{code}
  Program Definition head \{ l : list A | l <> 0 \} : A :=
    match l with
    | hd :: tl -> hd
    | [] -> _
    end.
\end{code}
Here the second branch can't be taken because we would have $l = []$ and
$l \neq 0$, but we may not have any placeholder element of type
$A$. There is also support for strucural and well-founded \verb|Fixpoint| definitions.

>>>>

The option type is the usual way to encode partiality in
\Coq. However we would more naturally constrain the argument \id{l} to
non-empty lists to be more faithful to the original \ML code, and also
attach a property to the result to express more precisely what \id{tail}
does:
\begin{code}
  Definition tail (l : list A) : (l <> []) -> 
  \{ l' : list A | exists a, a :: l' = l \}.
\end{code}
Now the definition's body requires some heavy plumbing of the code which is 
not affordable when programming. Moreover, the user is forced to give a
proof term justifying that \verb|l <> []| when calling the function.
We propose a solution to overcome these difficulties, based on the \ps{}
\cite{Shankar&Owre:WADT99} technique of \PVS{}
\cite{PVS-Semantics:TR}. It is separated into two phases. First, we have
a weak, \emph{decidable} type-checking
procedure which does not require proofs to be present in the code when
constructing objects of a subset type. In our new language \Russell, the
following is a well-typed term:
\begin{code}
  Definition tail ( l : list A | l <> [] ) : 
    \{ l' : list A | exists a, a :: l' = l \} := 
  match l with
    hd :: tl -> tl
  | [] -> []
  end
\end{code}

The specification shows that we are defining a partial function and
enforces a relation between input and output, using a dependent type, yet the
code remains as simple as the \ML{} definition. This is only possible
because we leave the proofs in both branches. After type-checking,
there is an automatic elaboration into partial \Coq terms, which collects
obligations the user has to prove. In our example,
the proof assistant will ask the user to prove that:
\begin{enumerate}
\item The list \verb|tl| has the property \verb|exists a, a :: tl = hd :: tl|
  and
\item In the context where
  \id{l} is a non-empty list and $\texttt{l} = []$, the list $[]$ has the property
  \verb|exists a, a :: [] = []| (which should be obvious as the context is contradictory).
\end{enumerate}

@@

In \Coq, the distinction between informative and propositional parts of
a term is formalized by the \Prop/\Set{} sorts. In \Russell, we have
special support for propositions appearing in subset types. By
delimiting the use of propositions, we can separate code from proof.

Informally, this idea is already present in mathematics. When you have
an element of subset $\{ x `: S `| P \}$, you can freely forget about the property $P$
and use any operation which is defined on $S$. Conversely, when you
want to use an operation defined on a subset, say $f : \{ x : `N `| P \} "->"
X$, you usually prove \emph{first} that you apply it only to elements 
having the desired property. For example $(f~2)$ is a correct application
only if $(P~2)$ is provable.
In the context of formal development of programs, such a workflow is not
entirely satisfactory because it forces one to create objects and prove
properties about them \emph{at the same time}. In \Coq{}, we are forced
to apply $f$ to an object $(\elts{2}{p})$ where $p$ is a proof of
$(P~2)$ (figure \ref{fig:cci-subset} presents the definition of the subset type in \Coq).
We would like to be able to prove that our usage of partial functions 
is correct only after the program is written. Subsets are particularly well-suited in this respect
because they separate the objects we want to manipulate and their associated
properties.

\begin{figure}[h]
  \begin{center}
    \def\type{\typed}
    \AXC{$`G \type A : \Set$}
    \AXC{$`G, x : A \type P : \Prop$}
    \LeftLabel{\irule{Subset}}
    \BIC{$`G \type \mysubset{x}{A}{P} : \Set$}
    \DP

@@

    \AXC{$`G \type a : A$}
    \AXC{$`G \type p : P[a/x]$}
    \AXC{$`G \type \mysubset{x}{A}{P} : \Set$}
    \LeftLabel{\irule{Element}}
    \TIC{$`G \type \elt{A}{(\lambda x : A.P)}{a}{p} : \mysubset{x}{A}{P}$}
    \DP
    \vspace{\infvspace}

    \AXC{$`G \type t : \mysubset{x}{A}{P}$}
    \LeftLabel{\irule{Subset-$\eltpit$}}
    \UIC{$`G \type \eltpit~t : A$}
    \DP
    \quad
    \AXC{$`G \type t : \mysubset{x}{A}{P}$}
    \LeftLabel{\irule{Subset-$\eltpip$}}
    \UIC{$`G \type \eltpip~t : P[\eltpit~t/x]$}
    \DP
  \end{center}
  \vspace{-2em}
  \caption{Subset type in \CIC}
  \label{fig:cci-subset} 
  \vspace{-1em}
\end{figure}

@@

%
\Russell{}'s type system is based on the Calculus of (Co-)Inductive
Constructions (figure \vref{fig:cic-decl-rules}) \cite{coquandhuet88},
with sigma types but without universes.  This restriction may be removed
in future work, but cause no problem for our main purpose which is
programming. We omit (co-)inductive constructs here, as they can be
considered as constants and leave their treatment as future work. The
judgement $`G \typed t : T$ means $t$ is a well-typed term of type $T$
in environment $`G$.

Following the presentation as a Pure Type System, the set of sorts
$\PTSsorts$ is defined as $\setproptype$. As usual, we let $s, s_i$ for
$i `: `N$ range over sorts. The axioms are $\PTSaxioms = \{ (\Set,
\Type), (\Prop, \Type) \}$ and the set of rules $\PTSrules$ is defined
by the functional relation $`A s_1 s_2, (s_1, s_2, s_2) `: \PTSrules$. We allow products $\Pi x :
A, B$ where $A : \Prop$ and $B : \Set$ but the user is encouraged to
encode them as $\Pi x : \{ \_ : unit `| A \}, B$ products to benefit from
the subtyping mechanism which we will define later.

@@

In \Russell, the conversion rule is replaced by a
new subsumption rule \irule{Subsum} (figure
\vref{fig:typing-decl-rules}) which will implement the subset
equivalence. The judgement $`G \typed T \subd U : s$ means $T$ is
equivalent to $U$ in environment $`G$, both sorted with $s$.
Subset equivalence will automatically insert first projections
(\irule{Subset-$\eltpit$}, \irule{SubSub}) and
element introductions (\irule{Element}, \irule{SubProof}), so we do not need
nor want these typing rules in \Russell. We do not want the second
projection to appear in \Russell terms either, because it would lead to
inconsistency. Indeed, in \PVS{} you can't manipulate proof terms, so you have no way to create a
term of type $0 \neq 0$ from the judgement $0 : \mysubset{x}{`N}{x \neq
  0}$, nor can you actually derive the latter, as it won't type-check.
If we allowed the second projection in \Russell, that would be doable.
Hence we leave only the subset type forming rule (figure
\vref{fig:typing-decl-rules}) in \Russell. With this restrictions, we
have the following property:
\begin{proposition}[Subset erasure and consistency]
  If we erase subset types from \Russell terms, leaving only the supports, we get a valid
  \CIC term, hence $`_$ is not provable in \Russell.
\end{proposition}

@@

\begin{paragraph}{}
We use a standard transformation to get an algorithm from our typing
rules, integrating the conversion (or subsumption) rule
(here \irule{Coerce}) to the premises of the other rules.
%We recall that $`G \type t : s$ is shorthand for $`G \type t : T$ and
%$T  `= s$ with $s `: \PTSsorts$ in a algorithmic presentation.
Our algorithmic judgement $`G \typea t : T$
(figure \vref{fig:typing-algo-rules})
reads: $t$ has type $T$ in environment $`G$.
We need to introduce the notion of support for subset types in order to
define our algorithmic system. Indeed, when typing an application, 
we need to ensure that the object we apply can be seen as an object of a
product type modulo the equivalence. 
% This notion is presented in \PVS{} by
% means of a $\mu_0$ operator which was readily adaptable in our setting
% (figure \vref{fig:mualgo-definition}). In Gang Chen's work, the notion is formalized by a
% new judgement $\Pi_{lub}$ which is essentialy equivalent to the operator
% $\mualgo$, except ours finds $\Sigma_{lub}$s too.
\begin{figure*}[t]
  \def\infvspace{0.5em}
    \def\type{\typea}
    \def\subt{\subta}
    \def\sub{\suba}
    \def\fCenter{\typea}
    \begin{center}
%     \WfEmpty\DP

@@

\end{paragraph}
\begin{paragraph}{}
  The next thing to do is to construct a decidable judgement for
  equivalence given two types. We denote it by $`G \typea T
  \suba U : s$ (figure \ref{fig:subtyping-algo-rules}).
  In our case, we have to check that conversion is decidable, which is the
  case. We also need to eliminate the transitivity and symmetry rules which are not
  syntax-directed. That is the purpose of the following theorem:
\begin{theorem}[Admissibility of transitivity and symmetry]
  \begin{enumerate}
  \item If $`G \typea T \suba U : s$ and $`G \typea U \suba V : s$, then $`G \typea T
    \suba V : s$.
  \item If $`G \typea T \suba U : s$ then $`G \typea U \suba T : s$.
  \end{enumerate}
\end{theorem}
\begin{proof}
  \begin{enumerate}
  \item By induction on the sum of the depths of the two derivations.
  \item By induction, using a thinning lemma.
  \end{enumerate}
\end{proof}
\end{paragraph}

@@

\begin{theorem}[Completeness]
  If $`G \typed t : T$ then there exists $T', s$ so that $`G \typea t : T'$ and $`G \typea T' \suba T : s$.
  If $`G \typed T \subd U : s$ then $`G \typea T \suba U : s$.
\end{theorem}

@@

We denote by $\typeafn{`G}{t}$ the function which returns the type of
term $t$ in context $`G$ such that $`G \typea t : \typeafn{`G}{t}$.

@@

We can now prove the following:
\begin{theorem}[Forgetful map correctness]
  If $`G \typec t : T$ then $`G^{\circ} \type t^{\circ} : T^{\circ}$ if $()^{\circ}$ is
  defined (ie, does not return terms containing $`_$) on $`G, t$ and $T$.
\end{theorem}

@@

We describe the case of application here, others are similar. First we
get the \Russell types $F$ and $U$ of the function $f$ and argument $u$ by calling the
typing function on both terms. Then we must ensure that $F$ can indeed
be seen as a product using the $\mualgo$ operator. We can then build
the coercion between $F$ and this product, and between $U$ and its
domain. We finally instantiate these coercions by their corresponding
interpreted objects and return their application.

@@

Similarly, we must add $\eta$ rules for all introduction terms of our
language if we want to be conservative and have unicity of
coercions. The later is an important property of our system, as it
corresponds in practice to the determinism of proof-obligations
generation. The complete equational theory of \CICq{} is given in figure
\vref{fig:eqccq}. It includes $\eta$, surjective pairing for dependent
sums and subsets ($\rho$) and proof-irrelevance for the second component
of subset objects.
Benjamin Werner has studied the addition of proof-irrelevance (rule
$\sigma$) in the Calculus of Constructions
\cite{Werner:ProofIrrelevance}. It is not a trivial extension and it has
far-reaching consequences on the model of the calculus
\cite{DBLP:conf/types/MiquelW02}. We direct the reader to these papers for further
information on the subject.
\begin{figure}[ht]
  \vspace{-1em}
  \ccqeqarr
  \vspace{-2em}
  \caption{Equational theory of \CICq}
  \label{fig:eqccq}
  \vspace{-1em}
\end{figure}

@@

\section{Implementation}
\label{sec:Program}
This mechanism has been implemented in \Coq and tested on simple
examples. We actually generate coercions simultaneously with typing 
in the implementation, which is sound thanks to the previous proofs. Our
typing is a clone of \Coq's original typing algorithm, so we benefit
from all the features of \Coq including implicit variables, notations and the existing
coercion system. The type-checker also transforms pattern-matching so
that an equality between the matched term and the pattern is present in
each branch typing context and it allows the user to put explicit holes in the term, for example:
\begin{code}
  Program Definition head ( l : list A | l <> 0 ) : A :=
    match l with
    | hd :: tl -> hd
    | [] -> _
    end.
\end{code}
Here the second branch can't be taken because we would have $l = []$ and
$l \neq 0$, but we may not have any placeholder element of type
$A$. There is also support for strucural and well-founded \verb|Fixpoint| definitions.
